---
title: "Data Science for Economics and Finance"
subtitle: "Lecture 5: Functions and parallel programming"
author:
  name: Dawie van Lill (dvanlill@sun.ac.za) | [Github](https://github.com/DawievLill)
  date: Lecture 5  "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_depth: 4
    toc_float: yes
always_allow_html: true
urlcolor: blue
mainfont: cochineal
sansfont: Fira Sans
monofont: Fira Code 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, dpi=300)
```
## References and packages

Most of the material that we will cover for the next few weeks is **heavily** influenced by (or directly copied from) the amazing lecture notes by Grant McDermott. You can visit his [Github page](https://github.com/uo-ec607/lectures) for more information.   

### R packages for this session

Run this piece of code to ensure that you have the appropriate packages installed for the session to follow. 

```{r, cache=F, message=F}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(pbapply, data.table, tidyverse)
```

## Return of the function

This first part of this section will be a recap of the lectures on programming from the first part of the course. We focus our attention on how to write functions and the related ideas in functional programming (especially as it applies to iteration). Most of these ideas overlap with what you were taught in the first few weeks, but there will be some components introduced that act as a continuation of what you covered before. The goal is to provide a solid foundation in these fundamental ideas, which can then lead us to our discussion on parallel programming in the final part of these notes. 

### Function syntax

You have already dealt with several functions in R. Some of them come pre-packaged with base R (e.g. `mean()`), while others are from external packages (e.g. `dplyr::filter()`). Functions in R all have the same basic syntax:

```r
function_name(ARGUMENTS)
```
Most of the time we use functions written by other people. However, as you know by now, you can write your own functions too. This is easy to do with the generic **`function()`** function. The syntax will again look familiar to you:

```r
function(ARGUMENTS) {
  OPERATIONS
  return(VALUE)
}
```

While we can write anonymous functions like the one above, we typically write functions because we want to reuse code. For this typical use-case it makes sense to name our functions. You can use either `=` or `<-` when assigning the function name. Try to give functions names that make sense to other people reading your code. 

```r
my_func =
  function(ARGUMENTS) {
    OPERATIONS
    return(VALUE)
  }
```

#### Function example

You should already have written some of your own functions, but as a quick reminder. Let us write a function that returns the cube of some input number.

```{r cube1}
cube =              ## Our function name
  function(x) {     ## The argument(s) that our function takes as an input
    x^3             ## The operation(s) that our function performs
  }
```

We can test whether this function works. Fingers crossed!

```{r cube 1a}
cube(3)
```

If my math is correct, it should be the case that $3^3$ is equal to $27$. So the function seems to be doing its job. In general we wouldn't write functions as simple as these, since they most likely already exist in base `R` (just look at the arithmetic functions in `R`) and will be more efficiently coded. However, in some cases we will need to write our own functions. 

In our example above we didn't specify the `return` value in the code block. This is because `R`, like many other programming languages, automatically returns the final object in that you created with the function. However, it is good practice to always assign return objects.

```{r cube2}
cube =              
  function(x) {     
    x_cub = x^3     ## Create an intermediary object (that will be returned)
    return(x_cub)   ## The value(s) or object(s) that we want returned.    
  }
```

Specifying a return value is also helpful when we want to return more than one object. For example, let's say that we want to remind our user what variable they used as an argument in our function:

```{r cube3}
cube = 
  function(x) {     ## The argument(s) that our function takes as an input
    x_cub = x^3     ## The operation(s) that our function performs
    return(list(value = x, value_cubed = x_cub)) ## The list of object(s) that we want returned.
  }
```

If we have multiple return objects we have to combine them in a list.  Remember that many objects in R contain multiple elements (vectors, data frames, and lists are all good examples of this). So we can also specify one of these "array"-type objects within the function itself if that provides a more convenient form of output. For example, we could combine the input and output values into a data frame:

```{r cube4}
cube = 
  function(x) { 
    x_cub = x^3 
    df = tibble(value = x, value_cubed = x_cub) ## Bundle up our input and output values into a convenient dataframe.
    return(df)
  }
```





























